---
title: "Book data set Analysis"
author: "Nabin B K / 30476867"
date: "2025-05-28"
output:
  word_document: default
  html_document: default
  pdf_document: default
always_allow_html: true
---

#1. Setup necessary library
```{r}
#import necessary library
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(httr)
library(tidyr)
```

#2. Load datasets (book, rating and user)

```{r }
#load data set file (csv) form the device
book_df <- read.csv("/Volumes/Drive S/college/lecture slides/data science/book recomendation/Books.csv")
rating_df <- read.csv("/Volumes/Drive S/college/lecture slides/data science/book recomendation/Ratings.csv")
user_df <- read.csv("/Volumes/Drive S/college/lecture slides/data science/book recomendation/Users.csv")
```

#3. display the datasets colmuns names 
```{r }
#display columns names
print(colnames(book_df))
```


```{r }
#display columns for rating dataset
print(colnames(rating_df))
```



```{r }
#display columns user dataset
print(colnames(user_df))
```

#4. show compact internal summary of datasets
```{r }
#disply structure of book data-frame using str() function
print(str(book_df))
```

```{r }
#disply structure of rating data-frame using str() function
print(str(rating_df))
```

```{r }
##disply structure of user data-frame using str() function
print(str(user_df))
```

#5. check null or missing value in datasets
```{r }
#check null value of book data-frame
print(sum(is.na(book_df)))
```

```{r }
#check null value of rating data-frame
print(sum(is.na(rating_df)))
```

```{r }
#check null value of user data-frame
print(sum(is.na(user_df)))
```


```{r }
#view null value inside user_df
print(str(user_df$Age))

```

```{r }
#remove this nuill value form Age cloumns
user_df <- user_df[!is.na(user_df$Age), ]
print(str(user_df$Age))
```

#6. confrim again null values presents or not

```{r }
#again check null value
print(sum(is.na(user_df)))
```


#7: Question number one: 
Q1. Write the code to calculate the average book rating for each publisher. en, tabulate the result for
the top 20 publishers (kable / kableextra would be desired). Using a bar chart, visualise the average
rating for the top 10 publishers with the most ratings. You can choose to use either RatingPGA or
RatingPGB, or both. Explain your findings.

```{r}
# merged datasets by their join key
merged_df <- merge(rating_df, user_df, by = "User.ID")
merged_df <- merge(merged_df, book_df, by = "ISBN")
# remove rows with missing age
merged_clean <- merged_df %>% filter(!is.na(Age), !is.na(Publisher), !is.na(Book.Rating))
# define RatingPGA (Age < 30)
RatingPGA <- merged_clean %>% filter(Age < 30)
# define RatingPGB (Age >= 30)
RatingPGB <- merged_clean %>% filter(Age >= 30)
# calculate average rating and count of ratings per publisher
publisher_stats <- RatingPGA %>%
  group_by(Publisher) %>%
  summarise(
    AvgRating = round(mean(Book.Rating, na.rm = TRUE), 2),
    RatingCount = n()
  ) %>%
  arrange(desc(AvgRating))
# top 20 publishers by average rating
top20_publishers <- publisher_stats %>% top_n(20, AvgRating)
# display as a styled HTML table
kable(top20_publishers, caption = "Top 20 Publishers by Average Rating")
```


# visualize the results using bar plot
```{r}
# top 10 publishers by number of ratings (for bar chart)
top10_by_count <- publisher_stats %>%
  slice_max(order_by = RatingCount, n = 10) #slice() sleect highest row with specific values
#create a ggplot object with 'top10_by_count' dataset
ggplot(top10_by_count, aes(x = reorder(Publisher, AvgRating), y = AvgRating, fill = Publisher)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = round(AvgRating, 2)), hjust = -0.1, size = 3.5, color = "black") +
  coord_flip() +  # flip x and y to make horizontal bars
  labs(title = "Top 10 Publishers with most ratings (RatingPGA)",
    x = "Publisher",
    y = "Average Rating"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",  # Remove legend (redundant)
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 10)
  ) +
  scale_y_continuous(limits = c(0, 10), expand = expansion(mult = c(0, 0.05))) #customzied the y-axis of the plot
```

# analysis of this part


#8 Question number two:
Analyse the distribution of book ratings by user age groupsfor example, ”0-18”, ”19-30”, ”31-45”, ”46-
60”, ”61+”. Visualise the distribution using a boxplot, with ‘Age‘ on the x-axis and ‘Book-Rating‘ on
the y-axis. You can choose to use either RatingPGA or RatingPGB or both. Explain your findings.
For this question, do NOT use tidyverse and only use the base R function.

# code according to the questions
```{r }
# Merge user and rating data
merged_df <- merge(rating_df, user_df, by = "User.ID")
# here create Age Groups
merged_df$AgeGroup <- cut( # Where cut() function divided numeric vectors in intevals
  merged_df$Age,
  breaks = c(-Inf, 18, 30, 45, 60, Inf),
  labels = c("0-18", "19-30", "31-45", "46-60", "61+"),
  right = TRUE
)
# it Display top 5 merged data frames.
print(head(merged_df))
```


```{r}
#according to table there exist NA value we need to remove this na value and fill with average one
#filter this one here
merged_df <- merged_df %>%
  filter(!is.na(Age), !is.na(AgeGroup))
# display top 5 rows after filtering
head(merged_df)
```

# Visualize the result in box plot with densy layers
```{r}
#generate random number
set.seed(123) 
# sample take 500 rows from merged_df (only rows with non-NA AgeGroup)
sampled_df <- merged_df[!is.na(merged_df$AgeGroup), ]
# Adjust sample dynamically
sample_size <- min(nrow(sampled_df), 500)  
#shuffling the rows and picking up random samples
sampled_df <- sampled_df[sample(nrow(sampled_df), sample_size), ]

# create a detailed boxplot with median, Q1, Q3, and outliers explicitly shown
ggplot(sampled_df, aes(x = AgeGroup, y = Book.Rating, fill = AgeGroup)) +
  geom_violin(alpha = 0.2) +  # Violin plot for smooth density visualization
  geom_boxplot(width = 0.2, fill = "lightblue", color = "#f1c21b", outlier.color = "#ff832b") +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "#198038") +
  labs(title = "Book Ratings Across Age Groups", x = "Age Group", y = "Book Rating") +
  theme_minimal()
```
```{r}
# Using your sampled_df or merged_df with AgeGroup and Book.Rating columns
ggplot(merged_df, aes(x = Book.Rating, color = AgeGroup, fill = AgeGroup)) +
  geom_density(alpha = 0.3) +  # Density plot with transparency
  labs(
    title = "Density Distribution of Book Ratings by Age Group",
    x = "Book Rating",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())

```

# tabulate the distributions of this results
```{R}
#Create counts of ratings per AgeGroup
age_counts <- table(merged_df$AgeGroup)
# Mean rating per age group
mean_ratings <- aggregate(Book.Rating ~ AgeGroup, data=merged_df, mean)
# Median rating per age group
median_ratings <- aggregate(Book.Rating ~ AgeGroup, data=merged_df, median)
# Standard deviation per age group
std_ratings <- aggregate(Book.Rating ~ AgeGroup, data=merged_df, sd)
# Min and Max per age group
min_ratings <- aggregate(Book.Rating ~ AgeGroup, data=merged_df, min)
max_ratings <- aggregate(Book.Rating ~ AgeGroup, data=merged_df, max)
# 25th and 75th percentiles per age group
quantile_25 <- tapply(merged_df$Book.Rating, merged_df$AgeGroup, function(x) quantile(x, 0.25))
quantile_75 <- tapply(merged_df$Book.Rating, merged_df$AgeGroup, function(x) quantile(x, 0.75))
# Combine into a data frame
distribution_df <- data.frame(
  AgeGroup = names(age_counts),
  Count = as.vector(age_counts),
  Mean = round(mean_ratings$Book.Rating, 2),
  Median = median_ratings$Book.Rating,
  StdDev = round(std_ratings$Book.Rating, 2),
  Min = min_ratings$Book.Rating,
  Q25 = quantile_25,
  Q75 = quantile_75,
  Max = max_ratings$Book.Rating
)
# display
print(distribution_df)
```

#9 Question Number Three:
Find out how the average book rating varies by location (country level) and tabulate them. Visualise
the top 5 countries with the highest average ratings using a bar chart for the 2 datasets. Discuss the
result. Is there any diﬀerence between the 2 datasets? (RatingPGA and RatingPGB)


#code according to question number three
```{r}
```

```{r}
# clean and extract country location (assuming "City, State, Country" format)
merged_df$Country <- trimws(sapply(strsplit(as.character(merged_df$Location), ","), tail, 1))
# define RatingPGA = users under 30
RatingPGA <- merged_df %>% filter(Age < 30 & !is.na(Age))
#display top 10 rows from RatingPGA
head(RatingPGA,10)
```

```{r}
# define RatingPGB = users 30 and above
RatingPGB <- merged_df %>% filter(Age >= 30 & !is.na(Age))
#display top 10 RatingPGB
head(RatingPGB,10)
```

# calulate average ratings 
```{r}
# Average ratings by country for RatingPGA
avg_rating_PGA <- RatingPGA %>%
  group_by(Country) %>%
  summarise(AvgRating = mean(Book.Rating, na.rm = TRUE), Count = n()) %>%
  filter(Count >= 10) #filter out countries with very few ratings
# Average ratings by country for RatingPGB
avg_rating_PGB <- RatingPGB %>%
  group_by(Country) %>%
  summarise(AvgRating = mean(Book.Rating, na.rm = TRUE), Count = n()) %>%
  filter(Count >= 10)
#print top 10 rows of average RatingPGA
head(avg_rating_PGA, 10)
```

# top five rows for ratingPGB
```{r}
#print top 10 rows of average RatingPGB
head(avg_rating_PGB, 10)
```

#plot top five countries
```{r}
# top 5 countries with highest average ratings
top5_PGA <- avg_rating_PGA %>% arrange(desc(AvgRating)) %>% head(5)
# Bar plot for RatingPGA
ggplot(top5_PGA, aes(x = reorder(Country, AvgRating), y = AvgRating)) +
  geom_bar(stat = "identity", fill = "orange") +
  geom_text(aes(label = round(AvgRating, 2)), 
            hjust = -0.2, color = "#198038", size = 4.5) +
  coord_flip() +
  labs(
    title = "Top 5 Countries - RatingPGA (Under 30)",
    x = "Location (Country)",
    y = "Average Rating"
  ) +
  theme_minimal() +
  ylim(0, max(top5_PGA$AvgRating) + 1)  # give space for label visibility

```
# Visualized the result based on RatingPGB
```{r}
top5_PGB <- avg_rating_PGB %>% arrange(desc(AvgRating)) %>% head(5)
# Bar plot for RatingPGB
ggplot(top5_PGB, aes(x = reorder(Country, AvgRating), y = AvgRating)) +
  geom_bar(stat = "identity", fill = "#fa4d56") +
  geom_text(aes(label = round(AvgRating, 2)), 
            hjust = -0.1, color = "#a56eff", size = 4.5) +
  coord_flip() +
  labs(
    title = "Top 5 Countries - RatingPGB (Above 30)",
    x = "Location (Country)",
    y = "Average Rating"
  ) +
  theme_minimal() +
  ylim(0, max(top5_PGA$AvgRating) + 1)  # Give space for label visibility

```

# Analysis of this part



#10: Question number Four:
Explore how book ratings vary by user age for books published aﬅer 2000. Use the data available to
investigate the relationship between user age and ratings, and present your findings in a bar chart
showing the average ratings for diﬀerent age groups. Compare the results of the two datasets.



# code aacording to the questions
```{r}
# Filter books published after 2000
# Merge with user info
merged_df <- merge(rating_df, user_df, by = "User.ID")
merged_df <- merge(merged_df, book_df[, c("ISBN", "Year.Of.Publication", "Publisher", "Book.Title")], by = "ISBN")
post2000_df <- merged_df %>%
  filter(Year.Of.Publication > 2000, !is.na(Age), !is.na(Book.Rating))
# Create AgeGroup column
post2000_df$AgeGroup <- cut(post2000_df$Age, 
                             breaks = c(0, 18, 25, 35, 50, 65, 100), 
                             labels = c("0-18", "19-25", "26-35", "36-50", "51-65", "66+"), 
                             right = FALSE)
set.seed(123)
#randomly slpit the dataset into two parts
random_split <- sample(1:nrow(post2000_df), size = 0.5 * nrow(post2000_df))
df_A <- post2000_df[random_split, ]
df_B <- post2000_df[-random_split, ]
#calculate average rating of per age group A
avg_rating_A <- df_A %>%
  group_by(AgeGroup) %>%
  summarise(AvgRating = mean(Book.Rating), Count = n())
#calculate average rating of per age group B
avg_rating_B <- df_B %>%
  group_by(AgeGroup) %>%
  summarise(AvgRating = mean(Book.Rating), Count = n())
print(avg_rating_A)
```

#print rating average RatingPGB
```{r}
print(avg_rating_B)
```


# visualize the results with remove NA values
```{r}
# Add labels to distinguish datasets
avg_rating_A$Dataset <- "Dataset A"
avg_rating_B$Dataset <- "Dataset B"
# Combine for plotting
combined_avg <- rbind(avg_rating_A, avg_rating_B)
#removed NA value
combined_avg <- na.omit(combined_avg)  # removes any rows with NA in any column
#plot the results
ggplot(combined_avg, aes(x = AgeGroup, y = AvgRating, fill = Dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.70), width = 0.5) +
  geom_text(aes(label = round(AvgRating, 2)),
            position = position_dodge(width = 0.70),
            vjust = -0.4,
            size = 3.5,
            color = "#005d5d") +
  scale_fill_manual(values = c("Dataset A" = "#f1c21b", "Dataset B" = "#ff832b")) +
  labs(
    title = "Average Book-Rating by Age Group (Books Published After 2000)",
    subtitle = "Comparison Between Two Dataset (A and B)",
    x = "Age Group",
    y = "Average Rating",
    fill = "Dataset: "
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 13),
    legend.position = "top",
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  ) +
  ylim(0, max(combined_avg$AvgRating, na.rm = TRUE) + 1)

```

# Analysis part






































